---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


# Weighted Average Longest Consonant Run
```{r}
longest_run <- rep(NA, nrow(babynames))
for (i in 1:nrow(babynames)) {
  name <- babynames$name[i]
  longest_run[i] <- max(sapply(str_split(name, "[aeiouAEIOU]+")[[1]], nchar))
}

consonant_run_df <- babynames %>%
  mutate(longest_run = longest_run) %>%
  group_by(year, sex) %>%
  summarize(weighted_longest_run = sum(n * longest_run)/sum(n), .groups = "drop")
```

```{r}
ggplot(aes(x = year, y = weighted_longest_run, color = sex), data = consonant_run_df) + 
  geom_line() + 
  scale_x_continuous(breaks = c(seq(1880, 2017, 10), 2017)) +
  scale_y_continuous(breaks = c(1.7, seq(1.75, 1.95, 0.05), 2.0), limits = c(1.7, 2)) + 
  scale_color_manual(values = c("purple2", "mediumaquamarine")) + 
  labs(x = "Year", y = "Number of Consonants", color = "Sex") + 
  ggtitle("Average max number of consonants in a row in first name, by year") + 
  theme_classic() + theme(legend.position = "bottom") 
```

```{r}
consonant_female <- filter(consonant_run_df, sex == "F")
consonant_male <- filter(consonant_run_df, sex == "M")
```

## FEMALE AVERAGE LONGEST CONSONANT RUN
```{r}
#code borrowed from slides, modified with try-catch 
aic_table2 <- function(data, P, Q) {
  table <- matrix(NA, (P+1), (Q + 1)) 
  for (p in 0:P) {
    for (q in 0:Q) {
      table[p+1, q+1] <- tryCatch({arima2::arima(data, order = c(p, 0, q))$aic}, error = function(e){NA})
    }
  }
  dimnames(table) <- list(paste("AR", 0:P, sep = ""),
                          paste("MA", 0:Q, sep = ""))
  return(table)
}
```

```{r}
female_table <- aic_table2(consonant_female$weighted_longest_run, 5, 5)
```
We get a warning about a convergence problem, and some of the p and q combinations return an error and display as NA in our table. We also not some mathematical impossibilities, where there is an increase larger than 2 units. (This is using the arima2 function though, so it's not that issue.)

Let's try fitting this with trend too:

```{r}
aic_table2_trend <- function(data, P, Q, d = 0, xreg) {
  table <- matrix(NA, (P+1), (Q + 1)) 
  for (p in 0:P) {
    for (q in 0:Q) {
      table[p+1, q+1] <- tryCatch({arima2::arima(data, order = c(p, d, q), xreg = xreg)$aic}, error = function(e){NA})
    }
  }
  dimnames(table) <- list(paste("AR", 0:P, sep = ""),
                          paste("MA", 0:Q, sep = ""))
  return(table)
}

female_table_trend <- aic_table2(consonant_female$weighted_longest_run, 5, 5)
```
Still getting a convergence issue, and AICs seem worse? 

```{r}
female_table_trend <- aic_table2_trend(consonant_female$weighted_longest_run, 5, 5, d = 0, xreg = consonant_female$year)
female_table_trend
```
```{r}
female_table_trend_d <- aic_table2_trend(consonant_female$weighted_longest_run, 5, 5, d = 1, xreg = consonant_female$year)
female_table_trend_d
```
Still getting convergence issues here. 

Let's try `auto.arima` just to see what it gives us?

```{r}
auto_model <- forecast::auto.arima(consonant_female$weighted_longest_run)
```
This suggest the best model is an ARIMA(1,2,1) model. Not sure what to do with the $d=2$ part? 

## MALE AVERAGE LONGEST CONSONANT RUN

```{r}
aic_table <- function(data, P, Q, D = 0) {
  table <- matrix(NA, (P+1), (Q + 1)) 
  for (p in 0:P) {
    for (q in 0:Q) {
      table[p+1, q+1] <- tryCatch({arima2::arima(data, order = c(p, D, q))$aic}, 
                                  error = function(e){
                                    print(e)
                                    NA})
    }
  }
  dimnames(table) <- list(paste("AR", 0:P, sep = ""),
                          paste("MA", 0:Q, sep = ""))
  return(table)
}

male_table <- aic_table(consonant_male$weighted_longest_run, P = 5,Q = 5)
male_table
```
A lot of problems here, most of these models have a "non-stationary AR part from CSS", which isn't suprising given the graph. 

Let's try this with $d=1$:
```{r}
male_tableD1 <- aic_table(consonant_male$weighted_longest_run, P = 5,Q = 5, D=1)
male_tableD1
```
This looks a lot better - we don't get any warnings or errors! This suggests our best model is an ARIMA(3,1,5) model. Let's check the roots of that model:

```{r}
male_consonant35 <- arima2::arima(consonant_male$weighted_longest_run, order = c(3, 1, 5))
male_consonant35
```
```{r}
plot(male_consonant35, order = c(3,1,5))
```
All of these inverse roots are inside the unit circle, which means this is causal and invertible. There are some roots that are right on the boundary, however. 
```{r}
male_consonant32 <- arima2::arima(consonant_male$weighted_longest_run, order = c(3, 1, 2))
plot(male_consonant32)
```
I tried playing around with decreasing P and Q to see if we can get all the roots off the boundary, but that is tricky. Try doing the profile likelihood thing like we did in class? 


# AVERAGE LENGTH

## FEMALE AVERAGE LENGTH

## MALE AVERAGE LENGTH

# AVERAGE NUMBER OF VOWELS 

## FEMALE AVERAGE NUMBER OF VOWELS

## MALE AVERAGE NUMBER OF VOWELS 

Sources:
1. Looked up what this error meant (https://stackoverflow.com/questions/7233288/non-stationary-seasonal-ar-part-from-css-error-in-r)
2. 
