---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}
library(stringr)
library(dplyr)
library(ggplot2)
```


```{r}
library(babynames)
head(babynames)
```

# AVERAGE LENGTH OF NAMES 
```{r}
length_df <- babynames %>%
  mutate(name_length = str_length(name)) %>%
  group_by(year, sex) %>%
  summarize(avg_length = mean(name_length)) 

ggplot(length_df, aes(x = year, y = avg_length, color = sex)) + geom_line()
```
This is the average length of a name, out of all names given. However, it doesn't take into consideration how many people were given a name. 

If we instead want the average length of someone's name, out of all people, it's this: 
```{r}
length_df2 <- babynames %>%
  mutate(name_length = str_length(name)) %>%
  mutate(name_length_people = name_length * n) %>%
  group_by(year, sex) %>%
  summarize(avg_length = sum(name_length_people)/sum(n)) 

ggplot(length_df2, aes(x = year, y = avg_length, color = sex)) + geom_line()
```


# AVERAGE NUMBER OF VOWELS
```{r}
vowel_df <- babynames %>%
  mutate(num_vowels = str_count(name, '[aeoiuAEOIU]')) %>%
  group_by(year, sex) %>%
  summarize(avg_num = mean(num_vowels)) 

ggplot(vowel_df, aes(x = year, y = avg_num, color = sex)) + geom_line()
```

Same thing as before! This is out of all names. If we instead we want it to be out of all people, we want:
```{r}
vowel_df2 <- babynames %>%
  mutate(num_vowels = str_count(name, '[aeoiuAEOIU]')) %>%
  mutate(num_vowels_people = num_vowels * n) %>%
  group_by(year, sex) %>%
  summarize(avg_num = sum(num_vowels_people)/sum(n)) 

ggplot(vowel_df2, aes(x = year, y = avg_num, color = sex)) + geom_line()
```


## Average number of constants in a row (across all people)
```{r}
max(sapply(str_split("Mia", "[!aeiouAEIOU]")[[1]], nchar))

longest_run <- rep(NA, nrow(babynames))
for (i in 1:nrow(babynames)) {
  name <- babynames$name[i]
  longest_run[i] <- max(sapply(str_split(name, "[!aeiouAEIOU]")[[1]], nchar))
}

consonant_run_df <- babynames %>%
  mutate(longest_run = longest_run) %>%
  group_by(year, sex) %>%
  summarize(weighted_longest_run = sum(n * longest_run)/sum(n)) 


ggplot(aes(x = year, y = weighted_longest_run, color = sex), data = consonant_run_df) + 
  geom_line() + 
  scale_x_continuous(breaks = c(seq(1880, 2017, 10), 2017)) +
  scale_y_continuous(breaks = c(1.7, seq(1.75, 1.95, 0.05), 2.0), limits = c(1.7, 2)) + 
  scale_color_manual(values = c("purple2", "mediumaquamarine")) + 
  labs(x = "Year", y = "Number of Consonants", color = "Sex") + 
  ggtitle("Average max number of consonants in a row in first name, by year") + 
  theme_classic() + theme(legend.position = "bottom") 
                                                                                                                                                          
```

# OLD IDEAS (DELETE UNLESS WE WANT THEM)

## Average number of "rare" letters - OLD 
```{r}
char_list <- strsplit(tolower(paste(babynames$name, collapse = "")), NULL)[[1]]
letter_count <- table(char_list) #note this is NOT weighted by how common a name is (which I think is okay?) -- we're taking rare here to mean it doens't show up in many names, not it isn't given to many people 
```
The five least common letters in names are x, q, w, p, and f. 




This is over all names: 
```{r}
rare_df <- babynames %>%
  mutate(num_rare = str_count(name, '[xqwpfXQWPF]')) %>%
  group_by(year, sex) %>%
  summarize(avg_num = mean(num_rare)) 

ggplot(rare_df, aes(x = year, y = avg_num, color = sex)) + geom_line()
```

This is over all people:
```{r}
rare_df2 <- babynames %>%
  mutate(num_rare = str_count(name, '[xqwpfXQWPF]')) %>%
  group_by(year, sex) %>%
  summarize(weighted_avg_num = sum(num_rare * n) / sum(n)) 

ggplot(rare_df2, aes(x = year, y = weighted_avg_num, color = sex)) + geom_line()
```

## Rarest three letter combinations - OLD 

```{r}
n <- nchar("Jaylin")
substrings <- sapply(1:(n-2), function(i) str_sub("Jaylin", i, i+2))



get_combo_dict <- function(df) {
  d <- c("test" = 0)
for (i in 1:nrow(df)) {
  name <- tolower(df$name[i])
  n <- nchar(name)
  substrings <- sapply(1:(n-2), function(i) str_sub(name, i, i+2))
  for (j in 1:length(substrings)) {
    #if in dictionary, increase count
    if (substrings[j] %in% names(d)) {
      d[substrings[j]] <- d[substrings[j]] + 1 
    }
    else{
      d[substrings[j]] <- 1 
    }
  }
}
return(d)
}

combo_dict <- get_combo_dict(babynames)

tail(combo_dict[order(-unlist(combo_dict))], 20) #these names must only appear once? (they'd duplicate if they were included in multiple years)
#should I have done this over all unique names, not all the rows in the dataset? 

#need to identify what names these go to? 
```







