---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r, cache = T}
longest_run <- rep(NA, nrow(babynames))
for (i in 1:nrow(babynames)) {
  name <- babynames$name[i]
  longest_run[i] <- max(sapply(str_split(name, "[!aeiouAEIOU]")[[1]], nchar))
}

consonant_run_df <- babynames %>%
  mutate(longest_run = longest_run) %>%
  group_by(year, sex) %>%
  summarize(weighted_longest_run = sum(n * longest_run)/sum(n)) 


ggplot(aes(x = year, y = weighted_longest_run, color = sex), data = consonant_run_df) + 
  geom_line() + 
  scale_x_continuous(breaks = c(seq(1880, 2017, 10), 2017)) +
  scale_y_continuous(breaks = c(1.7, seq(1.75, 1.95, 0.05), 2.0), limits = c(1.7, 2)) + 
  scale_color_manual(values = c("purple2", "mediumaquamarine")) + 
  labs(x = "Year", y = "Number of Consonants", color = "Sex") + 
  ggtitle("Average max number of consonants in a row in first name, by year") + 
  theme_classic() + theme(legend.position = "bottom") 
```


```{r}
consonant_female <- filter(consonant_run_df, sex == "F")
consonant_male <- filter(consonant_run_df, sex == "M")
```

AIC table code (modified from slides) with try-catch:
```{r}
#code borrowed from slides, modified with try-catch 
aic_table <- function(data, P, Q, D = 0, xreg = NULL) {
  table <- matrix(NA, (P+1), (Q + 1)) 
  for (p in 0:P) {
    for (q in 0:Q) {
      
      if (is.null(xreg)) {
        table[p+1, q+1] <- tryCatch({arima2::arima(data, order = c(p, D, q))$aic},
        error = function(e){
          print(c(p,q))
          print(e) 
          NA},
        warning = function(w) {
          print(c(p,q))
          print(w)
          NA
        })
      }
      else{
        table[p+1, q+1] <- tryCatch({arima2::arima(data, order = c(p, D, q), xreg = xreg)$aic},
        error = function(e){
          print(c(p,q))
          print(e) 
          NA},
        warning = function(w) {
          print(c(p,q))
          print(w)
          NA
        })
      }
    }
  }
  dimnames(table) <- list(paste("AR", 0:P, sep = ""),
                          paste("MA", 0:Q, sep = ""))
  return(table)
}


aic_table2 <- function(data, P, Q, d = 0, xreg) {
  table <- matrix(NA, (P+1), (Q + 1)) 
  for (p in 0:P) {
    for (q in 0:Q) {
      table[p+1, q+1] <- tryCatch({arima2::arima(data, order = c(p, d, q), xreg = xreg)$aic}, 
        error = function(e){
          print(c(p,q))
          print(e) 
          NA},
        warning = function(w) {
          print(c(p,q))
          print(w)
          NA
        })
    }
  }
  dimnames(table) <- list(paste("AR", 0:P, sep = ""),
                          paste("MA", 0:Q, sep = ""))
  return(table)
}

```


## Female Max Consonant Run

AIC table with $d = 0$ and no trend: 

```{r}
#d = 0 and no trend
consonant_female_0 <- aic_table(consonant_female$weighted_longest_run, P = 5, Q = 5, D = 0)

#d = 1 and no trend
consonant_female_1 <- aic_table(consonant_female$weighted_longest_run, P = 5, Q = 5, D = 1)

#d = 0 with trend
consonant_female_0_trend <- aic_table(consonant_female$weighted_longest_run, P = 5, Q = 5, D = 0, xreg = consonant_female$year)

#d = 1 with trend
consonant_female_1_trend <- aic_table(consonant_female$weighted_longest_run, P = 5, Q = 5, D = 1, xreg = consonant_female$year)
```
For our model with $d = 0$ and no trend, we get stationarity warnings for ARMA(3,2) and ARMA(5,4) and convergence issues with (1,4). For $d=1$ and no trend, we get a convergence issue at $4,4$. For $d=0$ with trend, we get a lot of stationarity and convergence issues. For $d = 1$ with trend, we get only a single convergence issue at $4,4$. 

For the table with $d=1$ and trend, our lowest AIC is ARIMA(3,1, 2), but an ARIMA(4,1, 2) is very close behind. For the table with $d= 1$ and no trend, the best model is a ARIMA(4,1,2). 

Let's do a likelihood ratio test between an ARIMA(4,1,2) with trend and one without trend. 

```{r}
female_consonant412 <- arima2::arima(consonant_female$weighted_longest_run, order = c(4, 1, 2))
female_consonant412_trend <- arima2::arima(consonant_female$weighted_longest_run, order = c(4, 1, 2), xreg = consonant_female$year)

female_consonant412
female_consonant412_trend
```
```{r}
1 - pchisq(2*(517.22- 517.21), df = 1) #this would reject our null that we need trend
```

But we actually don't get issues at all levels of the ARMA(p,q) model with no trend, so maybe we should be using this? 

```{r}
plot(arima2::arima(consonant_female$weighted_longest_run, order = c(4, 0, 3)))
```
This actually looks pretty good! 




## Male Max Consonant Run
